Hylomorphisms in Coq

### 2023-10-26 paper/paper.tex
Reshuffling the explanation and expand.

\maketitle

\section{Adjoint Folds}

\subsection{Canonical Adjoint Folds}
\[\begin{tikzcd}
	{\CatC} & \CatD
	\arrow[""{name=0, anchor=center, inner sep=0}, "L"', shift right=2, from=1-2, to=1-1]
	\arrow[""{name=1, anchor=center, inner sep=0}, "R"', shift right=2, from=1-1, to=1-2]
	\arrow["\dashv"{anchor=center, rotate=-90}, draw=none, from=0, to=1]
    \arrow["D"', loop, distance=2em, in=35, out=325, from=1-2, to=1-2]
    \arrow["LDR"', loop, distance=2em, in=215, out=145, from=1-1, to=1-1]
\end{tikzcd}
\]

@@

In the following section we instantiate the adjoint fold with some instances
taken from the literature.

\subsection{Example}
Say that we want to prove that the two functions below are well-defined

\begin{tabular}{LL}
  \operator{even} : \Nat \to \Bool                  &  \operator{odd} : \Nat \to \Bool\\
  \operator{even} (0) = \top                        &  \operator{odd}(0) = \top\\
  \operator{even} (n+1) = \neg \operator{odd}(n)    &  \operator{odd}(n+1) = \neg (\operator{even}(n))
\end{tabular}

In order to transform $\operator{even-odd}$ into two functions $\operator{even}$
and $\operator{odd}$ we are going to use the adjunction $\Delta \vdash \times$
where $\Delta : \CatC \to \CatC \times \CatC$ is the diagonal functor
$\Delta X = (X,X)$ and $\times : \CatC \times \CatC$ is the product seen as a
bifunctor

\[\begin{tikzcd}
	{\CatC \times \CatC} & \CatC
	\arrow[""{name=0, anchor=center, inner sep=0}, "\Delta"', shift right=2, from=1-2, to=1-1]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\times"', shift right=2, from=1-1, to=1-2]
	\arrow["\dashv"{anchor=center, rotate=-90}, draw=none, from=0, to=1]
    \arrow["D"', loop, distance=2em, in=35, out=325, from=1-2, to=1-2]
    \arrow["\Delta\circ D\circ \times "', loop, distance=2em, in=215, out=145, from=1-1, to=1-1]
\end{tikzcd}
\]
To get the natural isomorphism from this adjunction we can instantiate
(\ref{eq:adjoint-iso}) as follows
\[
  \floor{\cdot} : \CatC \times \CatC ((A, A) , (B_{1}, B_{2}) \iso \CatC (A , B_{1} \times B_{2}): \ceiling{\cdot}
\]
We can instantiate further the iso by setting $A = \Nat$, $B_{1} = \Bool$ and
$B_{2} = \Bool$.
\[
  \floor{\cdot}: \CatC \times \CatC ((\Nat, \Nat) , (\Bool, \Bool) \iso \CatC (\Nat , \Bool \times \Bool) : \ceiling{\cdot}
\]

To define this isomorphism we can just define the units of the adjunctions:
\begin{align*}
  & \eta_{A} : A \to A \times A\\
  & \eta_{A}= \pair{id_{A}}{id_{a}}\\
  & \epsilon_{(X,Y)} : ((X \times Y), (X \times Y) \to (X,Y)\\
  & \epsilon_{(X,Y)} = (\pi_{1}, \pi_{2})
\end{align*}
Notice that, in particular, the counit is given by the projections out of the
product.  Now, the right transponse $\ceiling{\cdot}$ of the adjunction can be
defined using the projections provided by the counit.  And in fact, the ceiling
can be defined as
\[
  \ceiling{g} = (\pi_{1},\pi_{2}) \cdot \Delta(g) =  (\pi_{1},\pi_{2}) \cdot (g,g) = (\pi_{1} \cdot g,\pi_{2}\cdot g)
\]
and the floor can be defined as
\[
  \floor{f}(a) = \pair{f(a)}{f(a)}
\]
At this point we are pretty much done as we can instantiate the diagrams
(\ref{eq:canonical-adjoint-fold}) where $D X = 1 + X$
\[
  \begin{tikzcd}
	{(1 + \Nat, 1 + \Nat)} && (1 + \Bool \times \Bool, 1 + \Bool \times \Bool) & { 1 + \Nat} & 1 + (\Bool \times \Bool) \\
	{(\Nat, \Nat)} && (\Bool, \Bool) & {\Nat} & \Bool \times \Bool
	\arrow["{\cata{a}}"', from=2-4, to=2-5]
	\arrow[""{name=0, anchor=center, inner sep=0}, "{\InIso}"', from=1-4, to=2-4]
	\arrow["{1 + \cata{a}}", from=1-4, to=1-5]
	\arrow["a", from=1-5, to=2-5]
	\arrow["{\ceiling{\cata{a}}}"', from=2-1, to=2-3]
	\arrow["{(1 +  \InIso,1 +  \InIso)}"', from=1-1, to=2-1]
	\arrow["{(1 + \cata{a}, 1 + \cata{a})}", from=1-1, to=1-3]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\ceiling{a}", from=1-3, to=2-3]
	\arrow[shorten <=35pt, shorten >=35pt, Rightarrow, 2tail reversed, from=1, to=0]
  \end{tikzcd}
\]
Now we define $a$ to be the following $D$ algebra:
\begin{align*}
  & a : 1 + \Bool \times \Bool \to \Bool\\
  & a (\operator{nil}) = (\top, \bot)\\
  & a (b_{1}, b_{2}) = (\neg b_{1}, \neg b_{2})
\end{align*}

It is easy to see that the catamorphism over the algebra $a$ equals
$\operator{even-odd}$ and by the adjoint fold property we know that its right
transpose satisfies the right hand side diagram. We now have to show that the
pair of functions $(\operator{even}, \operator{odd})$ equals the right transpose
of $\operator{even-odd}$, but this is easy to see by definition of the ceiling:
\[
  \ceiling{\operator{even-odd}} = (\pi_{1}\cdot \operator{even-odd}, \pi_{2}\cdot \operator{even-odd})
\]
and then it is an easy calculation to see that $\pi_{1}\cdot \operator{even-odd}$ equals $\operator{even}$ and
that $\pi_{2}\cdot \operator{even-odd}$ equals $\operator{odd}$.
\end{document}

>>>>

\maketitle
\newcommand{\comp}{\cdot}
\section{Introduction}
Say that we want to prove that the two functions below are well-defined

\begin{tabular}{LL}
  \operator{even} : \Nat \to \Bool                  &  \operator{odd} : \Nat \to \Bool\\
  \operator{even} (0) = \top                        &  \operator{odd}(0) = \bot\\
  \operator{even} (n+1) = \neg \operator{odd}(n)    &  \operator{odd}(n+1) = \neg (\operator{even}(n))
\end{tabular}

Now we recover the original two functions by setting
$\operator{even}' \eqdef \pi_{1} \comp \operator{even-odd}$
$\operator{odd}' \eqdef \pi_{2} \comp \operator{even-odd}$. The proof that these two pairs of functions are the same is an easy induction over the natural numbers.

This is an instance of a more general phenonema whose argumet uses category theory.
\section{Adjoint Folds}
In order to transform $\operator{even-odd}$ into two functions $\operator{even}$ and
$\operator{odd}$ we are going to use something called an ``adjunction''.
In particular, we have two functors. The diagonal functor $\Delta : \CatC \to \CatC \times \CatC$ defined as
$\Delta X = (X,X)$ and the product functor $\times : \CatC \times \CatC$. These two functors form an adjunction:
\[\begin{tikzcd}
	{\CatC \times \CatC} & \CatC
	\arrow[""{name=0, anchor=center, inner sep=0}, "\Delta"', shift right=2, from=1-2, to=1-1]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\times"', shift right=2, from=1-1, to=1-2]
	\arrow["\dashv"{anchor=center, rotate=-90}, draw=none, from=0, to=1]
    \arrow["D"', loop, distance=2em, in=35, out=325, from=1-2, to=1-2]
    \arrow["\Delta\circ D\circ \times "', loop, distance=2em, in=215, out=145, from=1-1, to=1-1]
\end{tikzcd}
\]
More formally, this means there is a one-to-one correspondence of maps, i.e. every map $(A,A) \to (B_{1}, B_{2})$ corresponds univocally to a map of type $A \to B_{1} \times B_{2}$. In the isomorphism below we use $\ceiling{\cdot}$ and $\floor{\cdot}$ to denote the maps witnessing the isomorphism called respectively the right and left transpose of the adjunction:
\[
  \floor{\cdot} : \CatC \times \CatC ((A, A) , (B_{1}, B_{2}) \iso \CatC (A , B_{1} \times B_{2}): \ceiling{\cdot}
\]
the right transponse $\ceiling{\cdot}$ of the adjunction can be
defined using the projections:
\[
  \ceiling{g} = (\pi_{1} \cdot g,\pi_{2}\cdot g)
\]
and the floor can be defined as using the pair into the product:
\[
  \floor{f}(a) = \pair{f(a)}{f(a)}
\]

We can instantiate further the iso by setting $A = \Nat$, $B_{1} = \Bool$ and
$B_{2} = \Bool$.
\[
  \floor{\cdot}: \CatC \times \CatC ((\Nat, \Nat) , (\Bool, \Bool) \iso \CatC (\Nat , \Bool \times \Bool) : \ceiling{\cdot}
\]
% At this point we are pretty much done as we can instantiate the diagrams
% (\ref{eq:canonical-adjoint-fold}) where $D X = 1 + X$
% \[
%   \begin{tikzcd}
% 	{(1 + \Nat, 1 + \Nat)} && (1 + \Bool \times \Bool, 1 + \Bool \times \Bool) & { 1 + \Nat} & 1 + (\Bool \times \Bool) \\
% 	{(\Nat, \Nat)} && (\Bool, \Bool) & {\Nat} & \Bool \times \Bool
% 	\arrow["{\cata{a}}"', from=2-4, to=2-5]
% 	\arrow[""{name=0, anchor=center, inner sep=0}, "{\InIso}"', from=1-4, to=2-4]
% 	\arrow["{1 + \cata{a}}", from=1-4, to=1-5]
% 	\arrow["a", from=1-5, to=2-5]
% 	\arrow["{\ceiling{\cata{a}}}"', from=2-1, to=2-3]
% 	\arrow["{(1 +  \InIso,1 +  \InIso)}"', from=1-1, to=2-1]
% 	\arrow["{(1 + \cata{a}, 1 + \cata{a})}", from=1-1, to=1-3]
% 	\arrow[""{name=1, anchor=center, inner sep=0}, "\ceiling{a}", from=1-3, to=2-3]
% 	\arrow[shorten <=35pt, shorten >=35pt, Rightarrow, 2tail reversed, from=1, to=0]
%   \end{tikzcd}
% \]
% Now we define $a$ to be the following $D$ algebra:
% \begin{align*}
%   & a : 1 + \Bool \times \Bool \to \Bool\\
%   & a (\operator{nil}) = (\top, \bot)\\
%   & a (b_{1}, b_{2}) = (\neg b_{1}, \neg b_{2})
% \end{align*}

We already seen that the catamorphism over the algebra $a$ equals
$\operator{even-odd}$. By using the right transpose we can derive the pair of original functions $(\operator{even}, \operator{odd})$:
\[
  \ceiling{\operator{even-odd}} = (\pi_{1}\cdot \operator{even-odd}, \pi_{2}\cdot \operator{even-odd})
\]
To show this pair of functions are the same as the original ones we appeal to
uniqueness of catamorphisms. Recall that catamorphisms for a generic $F$ algebra $a$ are the unique maps such that
\[
  \cata{a} = a \comp F(\cata{a})\comp \InIso^{\circ}
\]
This is represented on the right hand side of the following diagrams:
\[
  \begin{tikzcd}
	{(1 + \Nat, 1 + \Nat)} && (1 + \Bool^{2}, 1 + \Bool^{2}) & {1 + \Nat} & 1 + \Bool^{2} \\
	{(\Nat, \Nat)} && (\Bool, \Bool) & {\Nat} & \Bool^{2}
	\arrow["{\operator{eo}}"', dotted,  from=2-4, to=2-5]
	\arrow[""{name=0, anchor=center, inner sep=0}, "{\InIso}"', from=1-4, to=2-4]
	\arrow["{1 + \operator{eo}}", dotted, from=1-4, to=1-5]
	\arrow["a", from=1-5, to=2-5]
	\arrow["{\ceiling{\operator{eo}}}"', from=2-1, to=2-3]
	\arrow["{(1 +  \InIso,1 +  \InIso)}"', from=1-1, to=2-1]
	\arrow["{(1 + \operator{eo}, 1 + \operator{eo})}", from=1-1, to=1-3]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\ceiling{a}", from=1-3, to=2-3]
	\arrow[shorten <=20pt, shorten >=20pt, Rightarrow, 2tail reversed, from=1, to=0]
  \end{tikzcd}
\]
The double implication in the middle of these two diagrams states that the catamorphism on the right is in one-to-one correspondences with the right transpose of $\operator{eo}$ satisfying the right-hand side diagram. Notice the right-hand side is exactly the definition of the pair of functions $\operator{even}$ and $\operator{odd}$ and the diagram states these correspond to $\operator{even-odd}$ given by a catamorphism on the left. Since this latter is the unique catamorphism the one on the right is also unique such that it makes the right-hand side diagram commute.
\subsection{The General Case}

\[\begin{tikzcd}
	{\CatC} & \CatD
	\arrow[""{name=0, anchor=center, inner sep=0}, "L"', shift right=2, from=1-2, to=1-1]
	\arrow[""{name=1, anchor=center, inner sep=0}, "R"', shift right=2, from=1-1, to=1-2]
	\arrow["\dashv"{anchor=center, rotate=-90}, draw=none, from=0, to=1]
    \arrow["D"', loop, distance=2em, in=35, out=325, from=1-2, to=1-2]
    \arrow["LDR"', loop, distance=2em, in=215, out=145, from=1-1, to=1-1]
\end{tikzcd}
\]

@@

In the following section we instantiate the adjoint fold with some instances
taken from the literature.
\end{document}
